Efficiently Handling Guards in the low level design of RPython's tracing JIT

### 2012-07-23 talk/vmil2012/paper.tex
make david first author
add a list of the benchmarks used

\authorinfo{Carl Friedrich Bolz$^a$ \and David Schneider$^{a}$}
           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
           }
           {XXX emails}

@@

Although there are several publications about tracing jut-in-time compilers, to
our knowledge, there are none that describe the use and implementaiton of
guards in this context. With the following contributions we aim to sched some
light (to much?) on this topic.
The contributions of this paper are:
\begin{itemize}
 \item
\end{itemize}

@@

\begin{figure}
    \input{figures/log.tex}
    \caption{Optimized trace}
    \label{fig:trace-log}
\end{figure}
\section{Guards in the Backend}
\label{sec:Guards in the Backend}

After optimization the resulting trace is handed to the backend to be compiled
to machine code. The compilation phase consists of two passes over the lists of
instructions, a backwards pass to calculate live ranges of IR-level variables
and a forward one to emit the instructions. During the forward pass IR-level
variables are assigned to registers and stack locations by the register
allocator according to the requirements of the to be emitted instructions.
Eviction/spilling is performed based on the live range information collected in
the first pass. Each IR instruction is transformed into one or more machine
level instructions that implement the required semantics, operations withouth
side effects whose result is not used are not emitted. Guards instructions are
transformed into fast checks at the machine code level that verify the
corresponding condition.  In cases the value being checked by the guard is not
used anywhere else the guard and the operation producing the value can merged,
reducing even more the overhead of the guard. Figure \ref{fig:trace-compiled}
shows how an \texttt{int\_eq} operation followed by a guard that checks the
result of the operation are compiled to pseudo-assembler if the operation and
the guard are compiled separeated or if they are merged.

@@

Second a piece of code is generated for each guard that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline. In case the
condition checked in the guard fails execution and a side-exit should be taken
execution jumps the the trampoline. In the trampoline the pointer to the
\emph{low-level resume data} is loaded and jumps to generic bail-out handler
that is used to leave the compiled trace if case of a guard failure.

@@

As in previous sections the underlying idea for the design of guards is to have
a fast on trace profile and a potentially slow one in the bail-out case where
the execution takes one of the side exits due to a guard failure. At the same
time the data stored in the backend needed to rebuild the state should be be
as compact as possible to reduce the memory overhead produced by the large
number of guards\bivab{back this}.

As explained in previous sections, when a specific guard has failed often enogh
a new trace, refered to as bridge, starting from this guard is recorded and
compiled. When compiling bridges the goal is that future failures of the guards
that led to the compilation of the bridge should execute the bridge without
additional overhead, in particular it the failure of the guard should not lead
to leaving the compiled code prior to execution the code of the bridge.

The process of compiling a bridge is very similar to compiling a loop,
instructions and guards are processed in the same way as described above. The
main difference is the setup phase, when compiling a trace we start with a lean
slate. The compilation of a bridge is starts from a state (register and stack
bindings) that corresponds to the state during the compilation of the original
guard. To restore the state needed compile the bridge we use the encoded
representation created for the guard to rebuild the bindings from IR-variables
to stack locations and registers used in the register allocator.  With this
reconstruction all bindings are restored to the state as they were in the
original loop up to the guard.

Once the bridge has been compiled the trampoline method stub is redirected to
the code of the bridge. In future if the guard fails again it jumps to the code
compiled for the bridge instead of bailing out. Once the guard has been
compiled and attached to the loop the guard becomes just a point where
control-flow can split. The loop after the guard and the bridge are just
condional paths.

@@

>>>>

\authorinfo{David Schneider$^{a}$ \and Carl Friedrich Bolz$^a$}
           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
           }
           {XXX emails}

@@

Although there are several publications about tracing jut-in-time compilers, to
our knowledge, there are none that describe the use and implementation of
guards in this context. With the following contributions we aim to shed some
light (to much?) on this topic.
The contributions of this paper are:
\begin{itemize}
 \item
\end{itemize}

@@

\begin{figure}
    \input{figures/log.tex}
    \caption{Optimized trace}
    \label{fig:trace-log}
\end{figure}
% section Resume Data (end)
\section{Guards in the Backend}
\label{sec:Guards in the Backend}

After optimization the resulting trace is handed to the backend to be compiled
to machine code. The compilation phase consists of two passes over the lists of
instructions, a backwards pass to calculate live ranges of IR-level variables
and a forward one to emit the instructions. During the forward pass IR-level
variables are assigned to registers and stack locations by the register
allocator according to the requirements of the to be emitted instructions.
Eviction/spilling is performed based on the live range information collected in
the first pass. Each IR instruction is transformed into one or more machine
level instructions that implement the required semantics, operations withouth
side effects whose result is not used are not emitted. Guards instructions are
transformed into fast checks at the machine code level that verify the
corresponding condition.  In cases the value being checked by the guard is not
used anywhere else the guard and the operation producing the value can merged,
reducing even more the overhead of the guard. Figure \ref{fig:trace-compiled}
shows how an \texttt{int\_eq} operation followed by a guard that checks the
result of the operation are compiled to pseudo-assembler if the operation and
the guard are compiled separated or if they are merged.

@@

Second a piece of code is generated for each guard that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline. In case the
condition checked in the guard fails execution and a side-exit should be taken
execution jumps to the trampoline. In the trampoline the pointer to the
\emph{low-level resume data} is loaded and jumps to generic bail-out handler
that is used to leave the compiled trace in case of a guard failure.

@@

As in previous sections the underlying idea for the design of guards is to have
a fast on-trace profile and a potentially slow one in the bail-out case where
the execution takes one of the side exits due to a guard failure. At the same
time the data stored in the backend needed to rebuild the state should be be
as compact as possible to reduce the memory overhead produced by the large
number of guards\bivab{back this}.

As explained in previous sections, when a specific guard has failed often enough
a new trace, referred to as a \emph{bridge}, starting from this guard is recorded and
compiled. When compiling bridges the goal is that future failures of the guards
that led to the compilation of the bridge should execute the bridge without
additional overhead, in particular the failure of the guard should not lead
to leaving the compiled code prior to execution the code of the bridge.

The process of compiling a bridge is very similar to compiling a loop.
Instructions and guards are processed in the same way as described above. The
main difference is the setup phase. When compiling a trace we start with a clean
slate. The compilation of a bridge is started from a state (register and stack
bindings) that corresponds to the state during the compilation of the original
guard. To restore the state needed to compile the bridge we use the encoded
representation created for the guard to rebuild the bindings from IR-variables
to stack locations and registers used in the register allocator.  With this
reconstruction all bindings are restored to the state as they were in the
original loop up to the guard.

Once the bridge has been compiled the trampoline method stub is redirected to
the code of the bridge. In future if the guard fails again it jumps to the code
compiled for the bridge instead of bailing out. Once the guard has been
compiled and attached to the loop the guard becomes just a point where
control-flow can split. The loop after the guard and the bridge are just
conditional paths. \cfbolz{maybe add the unpatched and patched assembler of the trampoline as well?}

@@

\include{figures/benchmarks_table}
